### 2011-06-21 thesis/novelty-intro.tex
Text explaining the method and reasoning behind

>>>>

\section{Novelty Detection on the Conceptual Map}

This section presents now how to use graphical models produced by the conceptual map and the
novelty detection concepts introduced earlier on this chapter in order to detect room categories the
system is not aware of.

For that both models to approximate both the conditional and unconditional probability need to
be defined and then the ratio between them used to define a function over which a threshold
implements a novelty detection system.

\subsection{Approximating the Conditional Probability}
The actual semantic mapping of the environment happens using the \emph{conceptual map}
introduced on \autoref{sec:conceptual-map}. During it the agent instantiates a
\emph{chain graph} to model distribution of the sensed variables according to the known
concepts and categories introduced as hidden variables on the graph. Using that graphical model
the system is then able to propagate and find the most likely configuration of the represented
variables, for instance the semantic category of a specific room is modelled as an hidden variable
with states representing the semantic values, by calculating probabilities on that variable the
system finds its belief on the room belonging to a specific category.

Since the graphical model produced by the conceptual map tries to model the distribution of
variables assuming the knowledge of the agent holds true, it represents the way to model
$P(x|known)$, and allows to calculate the density probability that the given set of sensed
features $x$ are sensed given that all the variables and graph structure, including the category of
a specific room $a$ are correctly modelled by the agent knowledge.
This way the distribution modelled by a factor graph equivalent to the chain-graph used by
the conceptual map can be used as an approximation for the conditional probability of $x$.

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent] {$a$}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \node (shape1f) [factor] {}; \&
    \node (appearance1f) [factor] {}; \&
    \node (object1f) [factor] {}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

  \draw [-] (room1) -- (shape1f) -- (shape1l);
  \draw [-] (room1) -- (appearance1f) -- (appearance1l);
  \draw [-] (room1) -- (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);

  \draw [-] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}

\caption{Factor graph modelling the conditional probability distribution, case where the
         room category of room $a$ is known by the system.}
\end{figure}

\subsection{Approximating the Unconditional Probability}

With no knowledge on the unconditional probability the correct approach is to
assume a uniform distribution (\autoref{sec:max-entropy}).
That is represented in factor graphs by a graph without any factors: \autoref{fig:uniform-graph}.

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=20pt, row sep=10pt,ampersand replacement=\&] {
    \node (x1) [obs] {$S_1$}; \&
    \node (x2) [obs] {$A_1$}; \&
    \node (x3) [obs] {$O_1$}; \&
    \node (xi) [] {\dots}; \&
    \node (xn) [obs] {$X_n$}; \\
  };
\end{tikzpicture}
\caption{\label{fig:uniform-graph}A factor graph modelling a uniform
         distribution over the sensed set of features $x$.}
\end{figure}

Thought, very often there is extra knowledge that can be obtained about the distribution
of the variables that helps to model the unconditional distribution. With that, more accurate
models can be produced.
In this case, the system is interested in detecting only if a specific variable category is not
known. With that in mind, it was assumed that an unknown category can only influence variables
by using the same structure as the known variables. This is: it was assumed the structure of the
graph was correct even when a room has an unknown category.

\begin{figure}[h]
\centering
\subfloat[\label{fig:any-a}Since variable $a$ can be unknown, the agent does not has
knowledge on how to model the dashed relations.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \&
    \&
    \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,dashed] {$a$}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \nofactor {shape1f}{}{} {right=0pt}; \&
    \nofactor {appearance1f}{} {} {right=0pt}; \&
    \nofactor {object1f}{} {} {right=0pt}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };
  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);

  \draw [-,dashed] (room1) -- (shape1f);
  \draw [-,dashed] (room1) -- (appearance1f);
  \draw [-,dashed] (room1) -- (object1f);
  \draw [-,dashed] (room1) -- (r12f);
  \draw [-,dashed] (room1) -- (r13f);

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\qquad \qquad
\subfloat[Without knowledge on all states of $a$, all variables dependent on $a$ get directly
          dependent between each other, introducing a big single factor.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [factor] {}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \&
    \&
    \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

  \draw [-] (room1) -- (shape1l);
  \draw [-] (room1) -- (appearance1l);
  \draw [-] (room1) -- (object1l);
  \draw [-] (room2) -- (prop2);

  \draw [-] (room1) -- (room2);
  \draw [-] (room1) -- (room3);
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\caption{\label{fig:known-struct}Factor graph modelling distribution of sensed variables when variable $a$ can be unknown.}
\end{figure}

If the structure and all the other variables can be assumed to be correctly modelled by the
knowledge of the agent, then all variables that were directly dependent on $a$ become directly
dependent between them, introducing a single factor (\autoref{fig:known-struct}).
By approximating this factor the agent can then obtain a model for the unconditional probability.

In cases no other information is available the introduced factor can be considered uniform\footnote{Uniform factors do not influence the
distribution represented by the factor graph and so can be removed from the graph representation}
and dropped from the graph (see \autoref{fig:uniform-model}).
Nonetheless it may be the case that due the cost of labelling data, the agent does not has knowledge
on all the possible categories of $a$, but still has access to unlabelled data.
In that case unlabelled samples may be gathered and explored to achieve a better approximation of
the real distribution of variables.
Though the factor that needs to be approximated requires handling of all the connected variables
and enormous amounts of data may be needed to approximate it.

A simple approach can be performed by assuming the variables connected to the factor to be independent
but having a bias towards certain values. This approach is equivalent to connecting single factors
on each of variables and approximating those factors using unlabelled data as seen on \autoref{fig:independent-model}.

\begin{figure}[h]
\centering
\subfloat[\label{fig:uniform-model}Uniform Model]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,draw=none] {}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \&
    \&
    \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\qquad
\subfloat[\label{fig:independent-model}Independent Model] {
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,draw=none] {}; \& \&
    \node (room2) [latent] {}; \\
    \& \& \& \\
    \node (shape1f) [factor] {}; \&
    \node (appearance1f) [factor] {}; \&
    \node (object1f) [factor] {}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);

  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}

\caption{Factor graph modelling the conditional probability distribution, case where the
         room category of room $a$ is known by the system.}
\end{figure}

Other schemes may be tried to approximate the unconditional independent assuming that only
a specific variable is not necessarily known. For example modelling unconditional probability
with an hidden variable trained from the unlabelled data. Or by trying to split the single big
factor on factorizations of other factors.
Though the presented cases have interesting properties:

\begin{description}
\item[uniform model] - by assuming the factor to be uniform, it can be considered as not existent on
the graphical model and so special methods can be devised to calculate novelty on several variables
by analysing the messages that are propagated in and out of each variable when performing inference.

\item[independent model] - it can very easily be trained from unlabelled data. 
\end{description}

\subsection{Threshold}
After defining a model for the conditional and unconditional distributions, a threshold can be applied
over the ratio of those.

For that
